Proving Logical Atomicity using Lock Invariants\thanks{Supported by the National Science Foundation via grant \#1811894.}

### 2022-05-18 concurrency/ASL 2022/Verification of Concurrent Binary Search Tree.tex
slight edits

Later, the TaDA logic~\cite{tada}, which codified the notion of logical atomicity, presented an alternative set of lock specifications:
\[\langle b.\ (\mathsf{L}(\ell) \land \neg b) \vee (\mathsf{U}(\ell) \land b) \rangle\ \texttt{acquire}(\ell)\ \langle \mathsf{L}(\ell) \land b \rangle \qquad \langle \mathsf{L}(\ell) \rangle\ \texttt{release}(\ell)\ \langle \mathsf{U}(\ell) \rangle\]
An \texttt{acquire} operation may access the lock repeatedly, finding it in either the locked ($\mathsf{L}$) or unlocked ($\mathsf{U}$) state; if it sees the unlocked state ($b$ holds), then it can atomically move to a locked state and return. A release simply moves atomically from the locked state to the unlocked state. These operations make no mention of a protected resource \LB{maybe add this:\emph{, i.e.~they obey the pattern of atomic specs for $P_l = Q_l = True$,} ?}, but because they are atomic, they can interact with any global invariant (``shared region'' in TaDA's terminology): in particular, in combination with the invariant $(\mathsf{U}(\ell) * R) \vee \mathsf{L}(\ell)$, they can be used to derive the lock-invariant-based specs. It is also worth noting that these lock specifications are not axioms: they are proved for a simple spinlock implementation that uses atomic operations (e.g. compare-and-swap), which are the actual primitive operations in TaDA and the logics that follow it. In this style, tracking ownership of a lock is orthogonal to the lock assertion itself; a mechanism such as cancelable invariants~\cite{rustbelt-relaxed} can be used to both make the lock assertion $\mathsf{U}$/$\mathsf{L}$ publicly available and deallocate it once all threads are finished.

@@

\subsection{Atomicity with lock invariants}
The easiest way to synchronize a data structure is to add a lock that
each operation acquires at the start and releases at the end. These
\emph{coarse-grained} locking implementations are easy to verify with
invariant-style lock specs: the lock's invariant can contain the whole
data structure assertion (e.g. $\mathsf{BST}\:p\ t$), so 
%that whichever thread is currently executing an operation on the tree at
%$p$ has full ownership of the data structure. 
a thread always has full ownership of the data structure when
performing an operation on $p$.  If we only want to prove that the operation
is thread-safe, we might choose the lock invariant $\exists t.\
\mathsf{BST}\ p\ t$, asserting that we gain full ownership of the tree
in some state $t$ every time we acquire the lock. On the other hand,
if we want to prove specifications that describe changes to the data
structure---as in the atomic specification $\langle m.\ \treerep\
m\rangle\ \texttt{insert}(p, k, v)\ \langle \treerep\ (m[k \mapsto
v])\rangle$---we need ghost state to track the changes.

@@

Function \lstinline{pushdown_left} changes the structure of the tree in two ways. First, the rotation \lstinline{turn_left} (line 14) changes the ranges of the node to be deleted and its right child (e.g., nodes 40 and 55 in the first step of figure~\ref{delete-diagram}); this is a local change, as the ranges of all other nodes (even the children of the rearranged nodes) are preserved. This is not a linearization point, so the rotation must be considered not to change the abstract state of the tree. We accounted for this by quantifying over the abstract tree $t$ in the definition of $\treerep$; rearranging nodes without changing the key-value map still implements the same abstract state, and other threads cannot assume that the tree structure they have observed is still the current structure of the tree. Second, the final removal of the node (line 10) increases the ranges of the subtree below it; for example, in the last step of figure~\ref{delete-diagram}, the range of node 35 changes from $(30, 40)$ to $(30, 50)$ when the node is removed. This is the linearization point of the function. The change occurs while \lstinline{pushdown_left} holds only the lock of the deleted node, so the algebra for range ghost state must allow a node's range to increase at any time, even when the node's lock is held. Fortunately, \lstinline{insert} and \lstinline{lookup} only use ranges to assert that the target key is in the range of the current node, which is still true if the range increases.

@@

\section{Using the Specifications}
\begin{figure}[t]%[htp]
\begin{subfigure}[t]{0.4\textwidth}
\begin{lstlisting}[language = C, basicstyle=\small\ttfamily, numbers=left]
void* update_tree(void* t){
  a = 1;
  b = 2;
  c = 3; 
  d = 4;
  insert(t, 1, &a);
  insert(t, 2, &b);
  insert(t, 1, &c);
  insert(t, 2, &d);
  return NULL;
}
\end{lstlisting} 
\end{subfigure}\qquad
\begin{subfigure}[t]{0.55\textwidth}
 \begin{lstlisting}[language = C, basicstyle=\small\ttfamily]
int retrieve_value(treebox t){
  int y = -1;
  do {
     searchResult = lookup (t,  2);
     if (searchResult != NULL) {
        y = *((int *) searchResult);
     }
  } while (y != 4);
  void* searchResult = lookup (t,  1);
  return *((int *) searchResult);
}
\end{lstlisting}
\end{subfigure}
\caption{Client code}
\label{bst-conc-client}
\end{figure}
To demonstrate that the complexities of our lock-based proofs are hidden from prospective users, we verified a simple client program for the binary search tree, shown in figure \ref{bst-conc-client}. A producer thread running the \lstinline{update_tree} function populates a tree, while a consumer thread running \lstinline{retrieve_value} waits until key 2 is associated with a pointer to the value 4, and then returns the value associated with key 1. This is a data-structure-level version of the message-passing idiom, and demonstrates one of the advantages of atomic specifications: we can use them to synchronize threads and pass ownership of resources via data structure operations. The invariant for the BST \lstinline{t} is a simple state machine with three states, depending on the contents of the abstract map $m$: 1) $m(2) = \texttt{NULL}$, 2) $m(2) \mapsto i$ for some integer $i \neq 4$; and 3) $m(2) \mapsto 4 \land m(1) \mapsto 3$. In this final state, the invariant also holds the resources to be transferred from the producer to the consumer. We also include ghost state to reflect the fact that there is exactly one producer thread (so the producer always knows exactly which values are in the tree) and one consumer thread (so it can retrieve the resources without worrying about interference). The proof of the client uses only the atomic specifications of the BST operations, and is not affected by the style of lock specification.

>>>>

Later, the TaDA logic~\cite{tada}, which codified the notion of logical atomicity, presented an alternative set of lock specifications:
\[\langle b.\ (\mathsf{L}(\ell) \land \neg b) \vee (\mathsf{U}(\ell) \land b) \rangle\ \texttt{acquire}(\ell)\ \langle \mathsf{L}(\ell) \land b \rangle \qquad \langle \mathsf{L}(\ell) \rangle\ \texttt{release}(\ell)\ \langle \mathsf{U}(\ell) \rangle\]
An \texttt{acquire} operation may access the lock repeatedly, finding it in either the locked ($\mathsf{L}$) or unlocked ($\mathsf{U}$) state; if it sees the unlocked state ($b$ holds), then it can atomically move to a locked state and return. A release simply moves atomically from the locked state to the unlocked state. These operations do not specify what is protected by the lock, but because they are atomic, they can interact with any global invariant (``shared region'' in TaDA's terminology): in particular, in combination with the invariant $(\mathsf{U}(\ell) * R) \vee \mathsf{L}(\ell)$, they can be used to derive the lock-invariant-based specs. It is also worth noting that these lock specifications are not axioms: they are proved for a simple spinlock implementation that uses atomic operations (e.g. compare-and-swap), which are the actual primitive operations in TaDA and the logics that follow it. In this style, tracking ownership of a lock is orthogonal to the lock assertion itself; a mechanism such as cancelable invariants~\cite{rustbelt-relaxed} can be used to both make the lock assertion $\mathsf{U}$/$\mathsf{L}$ publicly available and deallocate it once all threads are finished.

@@

\subsection{Atomicity with lock invariants}
The easiest way to synchronize a data structure is to add a lock that
each operation acquires at the start and releases at the end. These
\emph{coarse-grained} locking implementations are easy to verify with
invariant-style lock specs: the lock's invariant can contain the whole
data structure assertion (e.g. $\mathsf{BST}\:p\ t$), so 
%that whichever thread is currently executing an operation on the tree at
%$p$ has full ownership of the data structure. 
that a thread always has full ownership of the data structure when
performing an operation.  If we only want to prove that the operation
is thread-safe, we might choose the lock invariant $\exists t.\
\mathsf{BST}\ p\ t$, asserting that we gain full ownership of the tree
in some state $t$ every time we acquire the lock. On the other hand,
if we want to prove specifications that describe changes to the data
structure---as in the atomic specification $\langle m.\ \treerep\
m\rangle\ \texttt{insert}(p, k, v)\ \langle \treerep\ (m[k \mapsto
v])\rangle$---we need ghost state to track the changes.

@@

The \lstinline{pushdown_left} function changes the structure of the tree in two ways. First, the rotation \lstinline{turn_left} (line 14) changes the ranges of the node to be deleted and its right child (e.g., nodes 40 and 55 in the first step of figure~\ref{delete-diagram}); this is a local change, as the ranges of all other nodes (even the children of the rearranged nodes) are preserved. This is not a linearization point, so the rotation must be considered not to change the abstract state of the tree. We accounted for this by quantifying over the abstract tree $t$ in the definition of $\treerep$; rearranging nodes without changing the key-value map still implements the same abstract state, and other threads cannot assume that the tree structure they have observed is still the current structure of the tree. Second, the final removal of the node (line 10) increases the ranges of the subtree below it; for example, in the last step of figure~\ref{delete-diagram}, the range of node 35 changes from $(30, 40)$ to $(30, 50)$ when the node is removed. This is the linearization point of the function. The change occurs while \lstinline{pushdown_left} holds only the lock of the deleted node, so the algebra for range ghost state must allow a node's range to increase at any time, even when the node's lock is held. Fortunately, \lstinline{insert} and \lstinline{lookup} only use ranges to assert that the target key is in the range of the current node, which is still true if the range increases.

@@

\section{Using the Specifications}
To demonstrate that the complexities of our lock-based proofs are hidden from prospective users, we verified a simple client program for the binary search tree, shown in figure \ref{bst-conc-client}. A producer thread running the \lstinline{update_tree} function populates a tree, while a consumer thread running \lstinline{retrieve_value} waits until key 2 is associated with a pointer to the value 4, and then returns the value associated with key 1. This is a data-structure-level version of the message-passing idiom, and demonstrates one of the advantages of atomic specifications: we can use them to synchronize threads and pass ownership of resources via data structure operations. The invariant for the BST \lstinline{t} is a simple state machine with three states, depending on the contents of the abstract map $m$: 1) $m(2) = \texttt{NULL}$, 2) $m(2) \mapsto i$ for some integer $i \neq 4$; and 3) $m(2) \mapsto 4 \land m(1) \mapsto 3$. In this final state, the invariant also holds the resources to be transferred from the producer to the consumer. We also include ghost state to reflect the fact that there is exactly one producer thread (so the producer always knows exactly which values are in the tree) and one consumer thread (so it can retrieve the resources without worrying about interference). The proof of the client uses only the atomic specifications of the BST operations, and is not affected by the style of lock specification.

\begin{figure}[htb]
\begin{subfigure}[t]{0.4\textwidth}
\begin{lstlisting}[language = C, basicstyle=\small\ttfamily, numbers=left]
void* update_tree(void* t){
  a = 1;
  b = 2;
  c = 3; 
  d = 4;
  insert(t, 1, &a);
  insert(t, 2, &b);
  insert(t, 1, &c);
  insert(t, 2, &d);
  return NULL;
}
\end{lstlisting} 
\end{subfigure}\qquad
\begin{subfigure}[t]{0.55\textwidth}
 \begin{lstlisting}[language = C, basicstyle=\small\ttfamily]
int retrieve_value(treebox t){
  int y = -1;
  do {
     searchResult = lookup (t,  2);
     if (searchResult != NULL) {
        y = *((int *) searchResult);
     }
  } while (y != 4);
  void* searchResult = lookup (t,  1);
  return *((int *) searchResult);
}
\end{lstlisting}
\end{subfigure}
\caption{Client code}
\label{bst-conc-client}
\end{figure}
